### 2022-03-10 symposium.tex
Submission ready

While the syntax used in this example would undoubtedly make some non-Scala programmers envious, its type safety leaves much to be desired.
First of all, the \emph{number} of variable bindings in the extractor is entirely opaque to Scala's type system, and left to the discretion of the programmer.
Furthermore, the values that come out of capturing groups can be null (when using optional captures) and thus require an additional layer of validation.
Those shortcomings might appear benign on small examples, but can easily turn into bugs when dealing with a large-scale code base.

@@

In this section, we give a brief introduction to two recent additions to Scala's type system: match types and generic tuples.
Our regular expression library makes extensive use those two features to enable type-safe and null-safe capturing group extraction.

@@

Match types provide first-class support for type-level computations by allowing programmers to perform pattern matching on types:

@@

\noindent
This new tuple representation is especially useful at the type level, where it allows programmers to manipulate tuples in a recursive manner.
For example, the following match type reverses the order of a tuple's elements:

@@

  \item The user interface, which ties everything together to compile, execute, and extract the results of a regular expression, while providing type-safe and null-safe experience.
\end{enumerate}

@@

Rest assured, this is not an oversight!
Regular expressions with trailing backslashes are invalid and result in runtime crashes (reported as an "unexpected internal error").
Even though detecting these types of errors at compile time is out of the scope of our library, doing so is still desirable.
If we invoke |Compile| on a regular expression with a trailing backslash, the compiler will run into a call to |charAt| with an out of bounds argument, catch the corresponding exception and report it as a compilation error, which is certainly better reporting than same error at run-time.

@@

\noindent
Similar to our handling of backslash characters, this implementation intentionally does not guard against out-of-bounds errors, since these errors correspond to ill-formed regular expressions.

@@

The typed-regex~\citep{akshay2021typed} Typescript library provides a type-safe API for extracting named capturing groups of regular expressions.
The implementation uses conditional types~\citep{microsoft2020typescript}, Typescript's type-level ternary operator, to statically analyze regular expressions.
Their approach, like ours, identifies optional capturing groups and marks them as such (using Typescript's optional properties).
At the time of writing, the library is still in an early stages of development and only supports a subset of the regular expression syntax, and notably lacks support for nested optional capturing group detection.

The safe-regex-rs~\citep{leonhard2021safe} Rust library takes a different approach towards the same goal: it uses Rust macros to code generate arity-specific extractors.
Unlike the
Their implementation is

Rust \url{https://gitlab.com/leonhard-llc/safe-regex-rs}

@@

In this \thesisOnly{chapter}\paperOnly{paper}, we introduce a new design for type-safe regular expressions in Scala.
We presented our type-level analysis, which identifies capturing groups and computes their nullability.
We built our library following the design of Scala's original regular expression API, in order to provide a frictionless migration path for programmers interested in the additional type safety.
We evaluated our implementation by running it against the QT3TS test suite and showed that, on those benchmarks, our type-level analysis only has a marginal impact on compilation times.

>>>>

While the syntax used in this example would undoubtedly make some non-Scala programmers envious, its type safety leaves much to be desired.
First of all, the \emph{number} of variable bindings in the extractor is entirely opaque to Scala's type system, and left to the discretion of the programmer.
Furthermore, the values that come out of capturing groups can be null (when using optional captures) and thus require an additional layer of validation.
Those shortcomings might appear benign on small examples, but can easily turn into bugs when dealing with a large-scale codebase.

@@

In this section, we give a brief introduction to two recent additions to Scala's type system: match types and generic tuples.
Our regular expression library makes extensive use of those two features to enable type-safe and null-safe capturing group extraction.

@@

Match types provide first-class support for type-level computations in the form of pattern matching on types:

@@

\noindent
This new tuple representation is especially useful at the type level, where it allows programmers to manipulate tuples recursively.
For example, the following match type reverses the order of a tuple's elements:

@@

  \item The user interface, which ties everything together to compile, execute, and extract the results of a regular expression, while providing a type-safe and null-safe experience.
\end{enumerate}

@@

Rest assured, this is not an oversight!
Regular expressions with trailing backslashes are invalid and result in runtime crashes (reported as an "unexpected internal error").
Even though detecting these types of errors at compile time is out of the scope of our library, doing so is still desirable.
If we invoke |Compile| on a regular expression with a trailing backslash, the compiler will run into a call to |charAt| with an out of bounds argument, catch the corresponding exception and report it as a compilation error, which is certainly better reporting that same error at run-time.

@@

\noindent
Similar to our handling of backslash characters, this implementation intentionally does not prevent out-of-bounds errors, since these errors correspond to ill-formed regular expressions.

@@

The typed-regex~\citep{akshay2021typed} Typescript library provides a type-safe API for extracting named capturing groups of regular expressions.
The implementation uses conditional types~\citep{microsoft2020typescript}, Typescript's type-level ternary operator, to statically analyze regular expressions.
Their approach, like ours, identifies optional capturing groups and marks them as such (using Typescript's optional properties).
At the time of writing, the library is still in the early stages of development and only supports a subset of the regular expression syntax, and notably lacks support for nested optional capturing group detection.

The safe-regex~\citep{leonhard2021safe} Rust library takes a different approach towards the same goal: it uses Rust macros to statically compile regular expressions and generate arity-specific extractors.
Performance-wise, macros improve over the status quo since they remove the need to compile regular expressions at runtime.
Safety-wise, code generation provides a straightforward solution to safe capturing group extraction.

@@

In this \thesisOnly{chapter}\paperOnly{paper}, we introduce a new design for type-safe regular expressions in Scala.
We presented our type-level analysis, which identifies capturing groups and computes their nullability.
We built our library following the design of the original Scala regular expression API, in order to provide a frictionless migration path for programmers interested in the additional type safety.
We evaluated our implementation by running it against the QT3TS test suite and showed that, on those benchmarks, our type-level analysis only has a marginal impact on compilation times.
